### 2022-10-04 AISTATS2023/paper.tex
Updates from Overleaf
changed the name

Recently, Machine Learning (ML) has flourished in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the various local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
for producing a concise global one is vulnerable to
misinterpretations. Under strong interactions and correlations between
features, the global explanation may obfuscate heterogeneous effects
that exist under the hood~\citep{herbinger2022repid}; a phenomenon
called aggregation bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental category
of global explainability methods. The objective of FE is to the
isolate and visualize the impact of a single feature on the
output.~\footnote{Often, FE methods also isolate the combined effect
  of a pair of features to the output. Combinations of more than two
  features are not usual, because they encounter, among others,
  visualization difficulties.} FE methods suffer from aggregation bias
because, often, the rationale behind the average effect is
unclear. For example, a feature with zero average effect may indicate
no effect on the output or, in contrast, highly positive in some cases
and highly negative in others. There are three widely-used FE methods;
Partial Dependence Plots (PDP)\citep{friedman2001greedy}, Marignal
Plots (MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method.

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition ALE does not inform the user about the level of
uncertainty, i.e., the heterogeneous effects hidden behing the average
effect due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be (partially) spotted by exploring the
Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation usign the (limited) samples of the training set,
ALE requires the additional \textit{bin-splitting} step. Bin-splitting
consists of partitioning the axis of the feature of interest in a
sequence of non-overlapping intervals (regions) and estimating a
constant effect in each one from the population of samples that fall
inside. Specifying an approriate sequence of regions is of particular
importance, since ALE's interpretation is meaningful only inside each
region \citep{molnar2022}. However, this crucial step has not raised
the appropriate attention, making the approximation vulnerable to
potential misinterpretations.

In this paper, we present Accumulated Regional Effects (ARE) an
extension of ALE for measuring the uncertainty of the explanation,
i.e.~how certain we are that the average explanation is valid if
applied to an instance drawn at random. We also provide an alternative
definition of ARE based on a sequence of constant-effect regions and
we define a metric of the dissimilarity of the two definitions. We
formally prove that under some conditions the dissimilarity is zero
and both definitions are equivalent. We, therefore, set up an
optimization problem where we search for the sequence of region that
minimizes the disimilarity, given enough training points iniside each
region for a robust estimation. Finally, we propose a
computationally-grounded algorithm for finding the optimal solution.

@@

\begin{itemize}
\item ARE, a FE method that extends ALE for quantifying the
  heterogeneous effects (uncertainty).
\item An interval-based definition of ARE and a formal proof that
  under some conditions it is equivalent to the initial definition.
\item A principled framework for automatically extracting regions with
  similar effects, improving the estimation and the interpretability
  of ALE plots.
\end{itemize}

@@

In this section, we describe the basic methods for FE and for
uncertainty quantification, focusing on ALE definition. We, then,
review the ALE approximation\citep{apley2020visualizing, gkolemis22},
describing some of its vulnerabilities.

@@

In real ML scenarios, the FE is estimated from the training set
examples. Therefore, \citep{apley2020visualizing} proposed dividing
the \(s\)-th axis in \(K\) bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that computes the local effects on the training instances using
auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

@@

\section{Accumulated Regional Effects (ARE)}
\label{sec:NAME-method}

ARE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting step for robust
estimations. At Section~\ref{sec:NAME-definition}, we define ARE and
at Section~\ref{sec:interval-based-estimation} we provide an
alternative definition of ARE based on variable-size bins, providing
important remarks and proofs for connecting it with the initial
definition. In Section~\ref{sec:interval-spliting}, we define the
problem of optimal bin-splitting and we propose an algorithmic
solution to the problem.

\subsection{ARE: ALE With Uncertainty Quantification}
\label{sec:NAME-definition}

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
ARE method formulates the effect at a specific point \(x_s\) with a
compact tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve
with a confidence region, i.e.
\(f^{\mathtt{ARE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

\subsection{Interval-Based Formulation and Approximation}
\label{sec:interval-based-estimation}

In real scenarios, the estimations are based on the limited instances
of the training set. Estimating \(\mu(z), \sigma(z)\) at the
granularity of a point is impossible, because the probability of
observing a sample inside the region \([x_s - h, x_s + h]\) tends to
zero, when \(h \to 0\). Therefore we define the regional-effect
\(\mu(z_1, z_2)\) and the regional-uncertainty \(\sigma(z_1, z_2)\),
as the summary statistics that characterize the effect and the
uncertainty inside the interval \([z_1, z_2)\):

@@

%
The regional-effect and the regional-uncertainty are the mean values
of the local effects and local uncertainty inside an interval. In
other words, they quantify what is the expected effect and the
expected uncertainty of a sample drawn at random from \([z_1, z_2]\)
given uniform distribution \(\mathcal{U}(z_1, z_2)\).

If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
redefine ARE using an interval-based formulation
\(\tilde{f}^{\mathtt{ARE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\subsection{Interval-Based Approximation}

The mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) can be directly estimated from
the set \(\mathcal{S}_k\) of dataset instances with the \(s\)-th
feature lying inside the interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):

@@

Under the assumption that the points are uniformly distributed inside
the intervals Eq.\eqref{eq:2} is an unbiased estimatior of
Eq.~(\ref{eq:mu_bin}). The same does not hold in the general case for
Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
is an unbiased estimator of the observable standard deviation
\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
is an overestimation of the real regional variance.

@@

\textit{If we define the residual \(\rho(z)\) as the difference
  between the expected effect at \(x_s\) and the regional effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then, the regional variance
  \(\sigma^2(z_1, z_2)\) equals to:}

@@

\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
\label{sec:interval-spliting}

@@

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of ARE method.

We split the simulation examples in two groups. The first group,
Section~\ref{sec:simulation-examples-1}, aims at showcasing that ARE
method is more accurate than PDP-ICE in quantifying the average effect
and the level of heterogeneous effects (uncertainty), when input
features are correlated. The second group,
Section~\ref{sec:simulation-examples-2}, illustrates that ARE method
achieves a better approximation (average effect and uncertainty) in
cases of limited samples, due to automatic bin-splitting. In both
groups, we choose to illustrate the most indicative examples; a more
extensive evaluation is provided in the Appendix.

@@

In this simulation, we will compare ARE method against PDP-ICE in
quantifying the main effect and the uncertainty. Since there is
ambiguity about the ground-truth first-order effect in cases of
correlated features,
e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
to the following piecewise linear function,

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
computed by ARE and PDP-ICE in three cases; (a) no interaction
(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we firstly compute the ground-truth average effect and
the uncertainty analytically (proofs in the Appendix) and then we
compare it against the approximation provided by each method. The
approximation is computed after generating \(N=300\) samples. As we
will see, despite the model's simplicity, PDP-ICE fail in quantifying
the effect correctly, due to correlation between features \(X_1\) and
\(X_2\). Finally, besides it is not the focus of the current examples,
we will also observe that ARE manages to correctly extract the
regions with constant effect. Details for obtaining the ground truth
effects and the experimental set-up are in the Appendix.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
  \caption{No interaction, Different weights: Feature effect for \(x_1\);
    Ground-truth vs (a) ARE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:ex-synth-1-2}
\end{figure}

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects in features \(x_1, x_3\), i.e., at any position
\(x_1\), the local effects dependend on the unknown value of \(X_3\)
and vice-versa. The effect of \(x_2\) continues to have
zero-uncertainty, since it does not appear in any interaction term. As
we observe in Figure~\ref{fig:ex-synth-1-3}, ARE captures perfectly
the effect of all features, and the uncertainty in all cases. As
expected, PDP-ICE quantify the effect and the uncertainty correctly
only in the case of \(x_3\), since \(X_3\) is independent from other
features. For the correlated features, i.e. \(x_1, x_2\), the average
effect computed by PDP and the uncertainty implied by ICE plots are
wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
  \caption{With interaction, equal weights: Feature effect for all
    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
    (a) ARE method at the left columns and (b) PDP-ICE at the right
    column.}
  \label{fig:ex-synth-1-3}
\end{figure}

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for seamlessly evaluating the competitive methods. Therefore, the main
purpose of this part is to demonstrate the usefulness of uncertainty
quantification in ARE, i.e., illustrate that modeling the
heterogeneous effects is important for the interpretation of an ALE
plot. For this reason, Secondly, we discuss about the FE plots obtaing with ARE and
PDP-ICE.

@@

For evaluating ARE in terms of bin-splitting we set-up the following
evaluation framework. We treat as ground-truth the effects computed on
the full training-set \(N=20000\) with dense fixed-size bin-splitting
(\(K=80\)). Given the big number of samples, we make the hypothesis
that the approximation with dense binning is close to the ground
truth. Afterwards we select less samples \(N=1000\) and we compare
ARE approximation with all possible fixed-size alternatives, in terms
of accuracy, i.e. we provide the quantities
\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma^2}, \mathcal{L}^{\rho^2}\).
\section{CONCLUSION}

>>>>

Recently, Machine Learning (ML) has flourished in critical domains,
such as healthcare and finance. In these areas, we need a combination
of accurate predictions along with meaningful explanations to support
them. For this reason there is an increased interest in Explainable AI
(XAI), the field that provides interpretations about the behavior of
complex black-box models. XAI literature distinguishes between local
and global explainability
techniques~\citep{Molnar2020interpretable}. Local methods explain a
specific prediction, whereas global methods explain the entire model
behavior. Global methods provide a universal explanation, summarizing
the various local explanations into a single interpretable outcome,
usually a number or a plot. If a user wants to get a rough overview
about which features are significant (feature importance) or whether a
particular feature has a positive or negative effect on the output
(feature effect), they should opt for a global explainability
technique. On the other hand, aggregating the individual explanations
is vulnerable to misinterpretations. Under strong interactions and
correlations between features, the global explanation may obfuscate
heterogeneous effects that exist under the
hood~\citep{herbinger2022repid}; a phenomenon called aggregation
bias~\citep{mehrabi2021survey}.

Feature effect (FE) \citep{Gromping2020MAEP} is a fundamental global
explainability category. The objective of FE is to isolate the impact
of a single feature on the output.~\footnote{Often, FE methods also
  isolate the combined effect of a pair of features to the
  output. Combinations of more than two features are not usual,
  because they encounter, among others, visualization difficulties.}
FE methods suffer from aggregation bias because, often, the rationale
behind the average effect is unclear. For example, a feature with zero
average effect may indicate no effect on the output or, in contrast,
highly positive in some cases and highly negative in others. There are
three widely-used FE methods; Partial Dependence Plots
(PDP)\citep{friedman2001greedy}, Marignal Plots
(MP)\citep{apley2020visualizing} and Aggregated Local Effects
(ALE)\citep{apley2020visualizing}. PDP and MP have been criticized for
computing erroneous effects when the input features are (highly)
correlated, which is a frequent scenario in many ML cases. Therefore,
ALE has been established as the state-of-the-art FE method.

However, ALE faces two crucial limitations, the first concerns ALE
definition and the second one ALE approximation. Regarding the
definition ALE does not inform the user about the uncertainty, i.e.,
the level of heterogeneous effects hidden behing the average effect
due to implicit feature interactions. In contrast, in PDPs the
heterogeneous effects can be (partially) spotted by exploring the
Individual Conditional Expectations
(ICE)\citep{goldstein2015peeking}. Regarding the approximation,
i.e.~the estimation usign the (limited) samples of the training set,
ALE requires the additional \textit{bin-splitting} step. Bin-splitting
consists of partitioning the axis of the feature of interest in a
sequence of non-overlapping intervals (regions) and estimating a
constant effect in each one from the population of samples that fall
inside. Specifying an approriate sequence of regions is of particular
importance, since ALE's interpretation is meaningful only inside each
region \citep{molnar2022}. However, this crucial step has not raised
the appropriate attention and the approximation is vulnerable to
potential misinterpretations.

In this paper, we present Uncertainty-awary Accumulated Local Effects
(UALE) an extension of ALE for measuring the uncertainty of the
explanation, i.e.~how certain we are that the average explanation is
valid if applied to an instance drawn at random. We show that choosing
an appropriate sequence of intervals is particularly important for a
robust estimation of the uncertainty, providing an algorithm that
automates this step. We also provide theoretical proofs for an
unbiased estimation of the uncertainty. We apply our method in multiple
synthetic and real world datasets for evaluating its performance.

@@

\begin{itemize}
\item UALE, a FE method that extends ALE for quantifying the
  heterogeneous effects (uncertainty).
\item A formal proof of the bin-splitting requirements for an unbiased
  approximation of the heterogeneous effects.
\item A framework for automatically extracting regions with similar
  effects, improving the estimation and the interpretability of ALE
  plots.
\end{itemize}

@@

In this section, we describe the basic methods for FE and uncertainty
quantification, focusing on ALE definition. We, then, review the ALE
approximation\citep{apley2020visualizing, gkolemis22}, describing some
of its vulnerabilities.

@@

In ML scenarios, ALE is estimated from the training set
examples. Therefore, \citep{apley2020visualizing} proposed dividing
the \(s\)-th axis in \(K\) bins and estimating the local effects in
each bin by evaluating the black box-function at the bin limits:
\begin{equation}
  \label{eq:ALE_accumulated_mean_est}
  \hat{f}^{\mathtt{ALE}}(x_s) = \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} \left [ f(z_{k}, \xc^i) - f(z_{k-1}, \xc^i)) \right ]
\end{equation}
We denote as \(k_x\) the index of the bin that \(x_s\) belongs to,
i.e. \(k_x: z_{k_x-1} \leq x_s < z_{k_x} \) and \(\mathcal{S}_k\) is
the set of training instance that lie in the \(k\)-th bin, i.e.
\( \mathcal{S}_k = \{ \xb^i : z_{k-1} \leq x^i_s < z_{k} \}
\). Afterwards,~\citep{gkolemis22} proposed the Differential ALE
(DALE) that computes the local effects on the training instances using
auto-differentiation:
\begin{equation}
  \label{eq:DALE_accumulated_mean_est}
  \hat{f}^{\mathtt{DALE}}(x_s) = \Delta x \sum_{k=1}^{k_x} \frac{1}{|\mathcal{S}_k|} \sum_{i:\mathbf{x}^i \in
    \mathcal{S}_k} f^s(\mathbf{x}^i)
\end{equation}
%
Their method has the advantages of remaining on-distribution even when
bins become wider and, most importantly, allows the recomputation of
the accumulated effect with different bin-splitting with near-zero
computational cost.

@@

\section{Uncertainty-Aware ALE (UALE)}
\label{sec:UALE}

UALE extends ALE for quantifying the level of heterogeneous effects
(uncertainty) and automates the bin-splitting. At
Section~\ref{sec:UALE-definition-1} we define UALE, at
Section~\ref{sec:UALE-definition-2} we redefine it using an
interval-based formulation, at
Section~\ref{sec:UALE-approximation} we provide an
approximation and at Section~\ref{sec:bin-spliting} we define and
solve the problem of optimal bin-splitting.

\subsection{UALE: ALE With Uncertainty Quantification}
\label{sec:UALE-definition-1}

We extend ALE definition of Eq.~(\ref{eq:ALE}) with a component for
quantifying the uncertainty. We denote as \(\mathcal{H}(z)\) the
uncertainty of the local effects at a specific point \(x_s=z\) and we
quantify it as the standard deviation of the local explanations:
\begin{equation}
  \label{eq:ALE_var}
  \mathcal{H}(z) := \sigma(z) = \sqrt{\mathbb{E}_{\Xcb|z}\left [ \left (\dfdx (z, X_c) - \mu(z) \right )^2 \right ] }
\end{equation}
\noindent
The uncertainty emerges from the natural characteristics of the
experiment, i.e.,~the feature correlations existent in the data
generating distribution and the implicit interactions of the black-box
function. We also define the accumulated uncertainty at \(x_s\), as
the accumulation of the standard deviation of the local effects along
the axis:
\begin{equation}
  \label{eq:ALE_acc_unc}
  f^{\mathtt{ALE}}_{\sigma}(x_s) = \int_{x_{s, min}}^{x_s} \sigma(z) \partial z
\end{equation}
\noindent
UALE formulates the effect at a specific point \(x_s\) with a compact
tuple that consists of the average effect and the uncertainty
\((\mu(z), \sigma(z))\) and visualizes them as a continuous curve with
a confidence region, i.e.
\(f^{\mathtt{UALE}}(x_s) := f^{\mathtt{ALE}}_{\mu}(x_s) \pm
f^{\mathtt{ALE}}_{\sigma}(x_s)\).

\subsection{Interval-Based Formulation and Approximation}
\label{sec:UALE-definition-2}

In real scenarios, all estimations are based on the instances of the
training set. It is impossible to estimate \(\mu(z), \sigma(z)\) at
the granularity of a point, because the probability of observing a
sample inside the region \([x_s - h, x_s + h]\) tends to zero, when
\(h \to 0\). Therefore we define the regional-effect \(\mu(z_1, z_2)\)
and the regional-uncertainty \(\sigma(z_1, z_2)\), as the summary
statistics that characterize the effect and the uncertainty inside the
interval \([z_1, z_2)\):

@@

%
Intuitively, the regional-effect and the regional-uncertainty are the
average effect \(\mu(z^*)\) and the uncertainty \(\sigma(z^*)\) that I
expect to get if draw a sample at random from the region
\([z_1, z_2]\) given a uniform distribution
\(z^* \sim \mathcal{U}(z_1, z_2)\).

If we also define as \(\mathcal{Z}_s\) the sequence of \(K+1\) points
that partition the \(s\)-th feature axis into \(K\) variable-size
intervals, i.e. \(\mathcal{Z}_s = \{z_0, \ldots, z_K\}\), we can
redefine UALE using an interval-based formulation
\(\tilde{f}^{\mathtt{UALE}}(x_s):= \tilde{f}^{\mathtt{ALE}}_{\mu}(x_s)
\pm \tilde{f}^{\mathtt{ALE}}_{\sigma}(x_s)\), where:

@@

\subsection{Interval-Based Approximation}
\label{sec:UALE-approximation}

For approximating the mean effect (Eq.~\eqref{eq:mu_bin}) and the
uncertainty (Eq.~\eqref{eq:var_bin}) we use the set \(\mathcal{S}_k\)
of dataset instances with the \(s\)-th feature lying inside the
interval
\( \mathcal{S}_k= \{ \mathbf{x}^i : z_{k-1} \leq x^i_s < z_k \} \):

@@

Eq.\eqref{eq:2} is an unbiased estimator of Eq.~(\ref{eq:mu_bin})
under the assumption that the points are uniformly distributed inside
the intervals. The same does not hold in the general case for
Eq.\eqref{eq:3}. In the general case, \(\hat{\sigma}^2(z_{k-1}, z_k)\)
is an unbiased estimator of the \textit{observable} standard deviation
\(\sigma_{obs}(z_1, z_2) = \sqrt{\frac{\int_{z_1}^{z_2}
    \mathbb{E}_{\xc|x_s=z} \left [ (f^s(z, \xc) - \mu(z_1, z_2) )^2
    \right] \partial z}{z_2 - z_1}}\), which as we prove in Theorem 1
is an overestimation of \(\sigma(z_1, z_2)\).

@@

\textit{If we define the residual \(\rho(z)\) as the difference
  between the expected effect at \(x_s\) and the regional effect, i.e
  \(\rho(z) = \mu(z) - \mu(z_1, z_2)\), then the regional variance
  \(\sigma^2(z_1, z_2)\) equals to:}

@@

\subsection{Bin-Splitting: Finding Regions With Homogeneous Effects}
\label{sec:bin-spliting}

@@

The simulation examples, where the data-generating distribution
\(p(\mathbf{X})\) and the predictive function \(f(\cdot)\) are defined
by us, enable the evaluation of competitive approaches against a solid
ground-truth. We follow this common XAI
practice~\citep{aas2021explaining, herbinger2022repid} for providing
secure empirical evidence about the superiority of UALE method.

We split the simulation examples in two groups. The first group,
Section~\ref{sec:simulation-examples-1}, aims at showcasing that UALE
method is more accurate than PDP-ICE in quantifying the average effect
and the level of heterogeneous effects (uncertainty), when input
features are correlated. The second group,
Section~\ref{sec:simulation-examples-2}, illustrates that UALE method
achieves a better approximation (average effect and uncertainty) in
cases of limited samples, due to automatic bin-splitting. In both
groups, we choose to illustrate the most indicative examples; a more
extensive evaluation is provided in the Appendix.

@@

In this simulation, we will compare UALE method against PDP-ICE in
quantifying the main effect and the uncertainty. Since there is
ambiguity about the ground-truth first-order effect in cases of
correlated features,
e.g.~\citep{apley2020visualizing,Gromping2020MAEP}, we limit ourselves
to the following piecewise linear function,

\begin{equation}
  \label{eq:synth-ex-1-function}
  f(\mathbf{x}) = \begin{cases}
                    f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $f_{\mathtt{lin}} < 0.5$ }\\
                    0.5 - f_{\mathtt{lin}} + \alpha f_{\mathtt{int}} & \text{if $0.5 \leq f_{\mathtt{lin}} < 1$}\\
                    \alpha f_{\mathtt{int}} &\text{otherwise}
                  \end{cases}
\end{equation}
%
where \(f_{\mathtt{lin}}(\mathbf{x}) = a_1 x_1 + a_2 x_2\) and
\(f_{int}(\mathbf{x}) = x_1x_3\). As we observe, the linear term
\(f_{\mathtt{lin}}\) includes the two correlated features and the term
\(f_{\mathtt{int}}\) interacts the two non-correlated variables. The
samples that we use in all examples are coming from the following
distribution: \(p(\mathbf{x}) = p(x_3)p(x_2|x_1)p(x_1)\) where
\(x_1 \sim \mathcal{U}(0,1)\), \(x_2 = x_1\) and
\(x_3 \sim \mathcal{N}(0, \sigma_3^2)\). We will test the effect
computed by UALE and PDP-ICE in three cases; (a) no interaction
(\(\alpha=0\)) and equal weights (\(a_1=a_2\)), (b) no interaction
(\(\alpha=0\)) and different weigths (\( a_1 \neq a_2 \)) and (c) with
interaction (\(\alpha \neq 0\)) with equal weights (\(a_1=a_2\)).

In all cases, we firstly compute the ground-truth average effect and
the uncertainty analytically (proofs in the Appendix) and then we
compare it against the approximation provided by each method. The
approximation is computed after generating \(N=300\) samples. As we
will see, despite the model's simplicity, PDP-ICE fail in quantifying
the effect correctly, due to correlation between features \(X_1\) and
\(X_2\). Finally, besides it is not the focus of the current examples,
we will also observe that UALE manages to correctly extract the
regions with constant effect. Details for obtaining the ground truth
effects and the experimental set-up are in the Appendix.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_1/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_1/pdp_ice_feat_0.pdf}
  \caption{No interaction, Equal weights: Feature effect for \(x_1\);
    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:synth-ex-1-case-1}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_2/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_2/pdp_ice_feat_0.pdf}
  \caption{No interaction, Different weights: Feature effect for \(x_1\);
    Ground-truth vs (a) UALE method at the left and (b) PDP-ICE at the
    right.}
  \label{fig:ex-synth-1-2}
\end{figure}

@@

In this case we activate the interaction term, i.e. \(a=1\), and we
set equal weights \(a_1=a_2=1\). The interaction term provokes
heterogeneous effects in features \(x_1, x_3\), i.e., at any position
\(x_1\), the local effects dependend on the unknown value of \(X_3\)
and vice-versa. The effect of \(x_2\) continues to have
zero-uncertainty, since it does not appear in any interaction term. As
we observe in Figure~\ref{fig:ex-synth-1-3}, UALE captures perfectly
the effect of all features, and the uncertainty in all cases. As
expected, PDP-ICE quantify the effect and the uncertainty correctly
only in the case of \(x_3\), since \(X_3\) is independent from other
features. For the correlated features, i.e. \(x_1, x_2\), the average
effect computed by PDP and the uncertainty implied by ICE plots are
wrong.

\begin{figure}[h]
  \centering
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_0.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_0.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_1.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_1.pdf}\\
  \includegraphics[width=.23\textwidth]{example_3/dale_feat_2.pdf}
  \includegraphics[width=.23\textwidth]{example_3/pdp_ice_feat_2.pdf}\\
  \caption{With interaction, equal weights: Feature effect for all
    features, \(x_1\) to \(x_3\) from top to bottom; Ground-truth vs
    (a) UALE method at the left columns and (b) PDP-ICE at the right
    column.}
  \label{fig:ex-synth-1-3}
\end{figure}

@@

In real-world datasets, it is infeasible to obtain the ground truth FE
for seamlessly evaluating the competitive methods. Therefore, the main
purpose of this part is to demonstrate the usefulness of uncertainty
quantification in UALE, i.e., illustrate that modeling the
heterogeneous effects is important for the interpretation of an ALE
plot. For this reason, Secondly, we discuss about the FE plots obtaing with UALE and
PDP-ICE.

@@

For evaluating UALE in terms of bin-splitting we set-up the following
evaluation framework. We treat as ground-truth the effects computed on
the full training-set \(N=20000\) with dense fixed-size bin-splitting
(\(K=80\)). Given the big number of samples, we make the hypothesis
that the approximation with dense binning is close to the ground
truth. Afterwards we select less samples \(N=1000\) and we compare
UALE approximation with all possible fixed-size alternatives, in terms
of accuracy, i.e. we provide the quantities
\(\mathcal{L}^{\mu}, \mathcal{L}^{\sigma^2}, \mathcal{L}^{\rho^2}\).
\section{CONCLUSION}
